---
title: "Exploratory Data Analysis Course Project2"
author: "Mary Li"
date: "November 12, 2015"
output: html_document
---

PM2.5 Emissions Data (summarySCC_PM25.rds): This file contains a data frame with all of the PM2.5 emissions data for 1999, 2002, 2005, and 2008. For each year, the table contains number of tons of PM2.5 emitted from a specific type of source for the entire year. Here are the first few rows.

fips: A five-digit number (represented as a string) indicating the U.S. county
SCC: The name of the source as indicated by a digit string (see source code classification table)
Pollutant: A string indicating the pollutant
Emissions: Amount of PM2.5 emitted, in tons
type: The type of source (point, non-point, on-road, or non-road)
year: The year of emissions recorded

Source Classification Code Table (Source_Classification_Code.rds): This table provides a mapping from the SCC digit strings in the Emissions table to the actual name of the PM2.5 source. The sources are categorized in a few different ways from more general to more specific and you may choose to explore whatever categories you think are most useful. For example, source “10100101” is known as “Ext Comb /Electric Gen /Anthracite Coal /Pulverized Coal”.

You can read each of the two files using the readRDS() function in R. For example, reading in each file can be done with the following code:

```{r, echo=FALSE, mesage=FALSE, warning=FALSE}

NEI <- readRDS("./exdata-data-NEI_data/summarySCC_PM25.rds")
SCC <- readRDS("./exdata-data-NEI_data/Source_Classification_Code.rds")

```
###### Question #1.
* Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.

```{r, echo=FALSE, mesage=FALSE, warning=FALSE}

library(dplyr)
by_year <- group_by(NEI, year)
emmision_by_year <- summarize(by_year,
  totalEmmision = sum(Emissions)
)
plot(emmision_by_year, type="l",xlab = "Year",  main = "Total Emissions in the United States",    ylab = "Total PM2.5 Emissions" )


# # Sum total emissions by year
# emissions <-   aggregate(Emissions ~ year,NEI, sum)
# # Generate a plot of data 
# barplot(emissions$Emissions, names.arg=emissions$year, xlab = "Year",  main = "Total Emissions in the United States",    ylab = "Total PM2.5 Emissions" )
## Export image as png
## Note did not specify the  width and height as the default values 480 
# dev.copy(png, file = "plot2.png")
# dev.off()


```
###### Question #2.
* Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008? Use the base plotting system to make a plot answering this question.

```{r, echo=FALSE, mesage=FALSE, warning=FALSE}

library(dplyr)
Baltimore_by_year <- group_by(filter(NEI, fips=="24510"), year)
emmision_by_year <- summarize(Baltimore_by_year,
  totalEmmision = sum(Emissions)
)

# png("plot2.png", width=480, height=480)
plot(emmision_by_year, type="l",main="Baltimore total PM2.5 emission over years", xlab="Year",  ylab = "Total PM2.5 Emissions" )
# dev.off()

# # Sum Baltimore emissions data by year
# baltimoreEmissions <- aggregate(Emissions ~ year, NEI[NEI$fips=="24510",],sum)
# # Generate a plot of data 
# barplot(baltimoreEmissions$Emissions, names.arg=baltimoreEmissions$year, xlab = "Year",  main = "Total Emissions in Baltimore",    ylab = "Total PM2.5 Emissions" )


```

###### Question #3.
* Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008? Use the ggplot2 plotting system to make a plot answer this question.

```{r, echo=FALSE, mesage=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
Baltimore_by_year <- group_by(filter(NEI, fips=="24510"), year, type)
emmision_by_year <- summarize(Baltimore_by_year,
  totalEmmision = sum(Emissions)
)
# png("plot3.png", width=480, height=480)
qplot(year, totalEmmision, data = emmision_by_year, facets = .~ type, geom =c("point", "smooth"))
# dev.off()

```

###### Question #4 
* Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?


```{r, echo=FALSE, mesage=FALSE, warning=FALSE}

combined_data <- inner_join(NEI, SCC, by=c("SCC"="SCC"))

# by looking at the actual text of the SCC.Level columns, and some initial assesement on how to extract the key words,
# I decided to use SCC.Level.One for combustion and SCC.Level.Three for coal to find 
# the source that's coal combustion-related records
coal_combustion_L1 <- grep("combustion", combined_data$SCC.Level.One, ignore.case = TRUE )
coal_combustion_L3 <- grep("coal", combined_data$SCC.Level.Three, ignore.case = TRUE )

coal_combustion <- intersect(coal_combustion_L1,coal_combustion_L3)

US_Emissions_by_year <- group_by(combined_data[coal_combustion,], year)
US_by_year <- summarize(US_Emissions_by_year, totalEmmision = sum(Emissions))

# png("plot4.png", width=480, height=480)
p <- qplot(year, totalEmmision, data = US_by_year, geom =c("point", "smooth"))
p + ggtitle("United States Coal Combustion-related emissions over years")
# dev.off()

# 
# # selecting coal combustion related data
# cl   <- grepl("coal", SCC$EI.Sector, ignore.case=TRUE)
# sc   <- SCC[cl,]$SCC
# coal <- subset(NEI, NEI$SCC %in% sc)
# q <- as.data.table( aggregate(coal$Emissions, by=list(coal$year), FUN=sum, na.rm=TRUE) )
# setnames(q, c("Year", "Emissions") )
# 
# png("plot4.png", width=512, height=512)
# plot(q$Year, q$Emissions/1.0e+3,
#      type="o", pch = 19, col=c("red"),
#      xlab="Year", ylab="PM2.5 Emissions, thou tons",
#      ylim=c(300, 600),
#      main="Total PM2.5 Emissions from Coal Combustion, 1999-2008")
# dev.off()


```

###### Question #5 
* How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?

```{r, echo=FALSE, mesage=FALSE, warning=FALSE}

combined_data <- inner_join(NEI, SCC, by=c("SCC"="SCC"))

Baltimore_by_year <- group_by(filter(combined_data, fips=="24510" & Data.Category=="Onroad"), year)
Baltimore_emmision_by_year <- summarize(Baltimore_by_year,
                                        totalEmmision = sum(Emissions)
)

png("plot5.png", width=480, height=480)
p <- qplot(year, totalEmmision, data = Baltimore_emmision_by_year, geom =c("point", "smooth"))
p + ggtitle("Baltimore Motor vehicle emissions over years")
dev.off()

```

###### Question #6 
* Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in
Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in 
motor vehicle emissions?


```{r, echo=FALSE, mesage=FALSE, warning=FALSE}

combined_data <- inner_join(NEI, SCC, by=c("SCC"="SCC"))

library(dplyr)
Baltimore_by_year <- group_by(filter(combined_data, fips=="24510" & Data.Category=="Onroad"), year)
Baltimore_emmision_by_year <- summarize(Baltimore_by_year,
                                        Baltimore_totalEmmision = sum(Emissions)
)

LosAngeles_by_year <- group_by(filter(combined_data, fips=="06037" & Data.Category=="Onroad"), year)
LosAngeles_emmision_by_year <- summarize(LosAngeles_by_year,
                                         LosAngeles_totalEmmision = sum(Emissions)
)

joined_data <- inner_join ( Baltimore_emmision_by_year,LosAngeles_emmision_by_year)

library(reshape2)

# organize the data in a desirable way for further use
Yr <- rep(joined_data$year,2)
df_melted <- melt(joined_data[2:3])
df_melted <-cbind(Yr,df_melted)

# calculate the rate of change in percentage for both cities
df_melted_changes <- mutate(df_melted, pct_changes=c(0.0, (df_melted$value[2:4]-df_melted$value[1:3])/df_melted$value[2:4]*100,
                            0.0, (df_melted$value[6:8]-df_melted$value[5:7])/df_melted$value[5:7]*100))

# this plot shows the overall change by Boxplot method, so we can see the scale of the changes over all
# we see the magnitude of changes are much higher in Baltimore than Los Angeless even though the absolute 
# value of total emission per year is way higher in Los Angeles than Baltimore
pct_Changes <- ggplot(df_melted_changes, aes(x=Yr, y=pct_changes, color=variable)) +
  geom_boxplot()+
  ggtitle("Motor vehicle emissions Overall Changes(%) ") +
  theme(legend.position = "bottom") +
  labs( x = "Year", y = "% changed") 

# This plot shows the emissions per year for both cities, here we do not see the scale of changes as well
# compare to the boxplot 
abslute_Changes <- ggplot(df_melted_changes, aes(x=Yr, y=value, color=Yr)) +
   geom_point(aes(fill=Yr))+
  facet_grid( ~ variable) +
  ggtitle("Motor vehicle total emissions over years") #+

# Put two plots together to convey the analysis results
# library("gridExtra")
png("plot6.png", width=1000, height=400)
grid.arrange(abslute_Changes, pct_Changes, ncol=2)
# dev.off()


```




